#load libraries
```{r}
library(geomorph)
library(abind)
library(rgl)
library(rlang)
library(ggplot2)
library(vegan)
library(Morpho)
library(dplyr)
library(stats)
library(ape)
library(picante)
library(phytools)

#set working directory
setwd("C:/Github/Dasyurid-skull-shape")
# setwd("C:/Users/20063747/Dropbox/Research/Dasyuridae_skull_evolution/2023")
```

#Import Landmark data, wireframe and classifiers
```{r}
Rawdata<-read.morphologika('Dasyurid_LMdata.txt')

Wireframe <- read.csv("Wireframe.csv")

#import classifier file csv
Rawclassifier<-read.csv('classifier.csv')
```

#Generalised Procrustes analyses
```{r}
procrustes<-gpagen(Rawdata$coords)

plotAllSpecimens(procrustes$coords)
```

#PCA of all individuals
```{r}
PCA_data <- gm.prcomp(procrustes$coords)
summary(PCA_data)
#PC1 23.5, PC2 16.8

PCA_data$d #eigenvalues (not standardised)
sum(PCA_data$d)
EVALUES <- 100*(PCA_data$d)/sum(PCA_data$d) 
plot(EVALUES)   #scree plot
```

#PCA graph of all individuals coloured by diet 
```{r}
diet <- as.factor(Rawclassifier$Diet) 
color1= rep(NA, length=length(Rawclassifier$DIET))
color1[which(Rawclassifier$Diet=="1")]= "brown"
color1[which(Rawclassifier$Diet=="2")]= "green"
color1[which(Rawclassifier$Diet=="3")]= "yellow"
color1[which(Rawclassifier$Diet=="4")]= "blue"
color1[which(Rawclassifier$Diet=="5")]= "red"

PCA_All_data <- plot(PCA_data$x[,1], PCA_data$x[,2], bg = color1, col= "black", pch = 21, cex = 2.5, 
                     asp = F, xlab="Principal Component 1 (23.5%)", ylab="Principal Component 2 (16.8%)")
```

#AVERAGE PER SPECIES FROM HERE ON
#Make a mean set of landmarks per species
```{r}
group <- factor(Rawclassifier$Species1)
levels(group)
species_coords <- coords.subset(A=procrustes$coords, group= group)
names(species_coords)
group_means <- lapply(species_coords, mshape)
mean_skull <- (aggregate(two.d.array(procrustes$coords)~ group, FUN = mean))[-1]
rownames(mean_skull) <- levels(group)
mean_skull <- arrayspecs(mean_skull,35,3)

#Visualise the first animal in this list (An.flav) against the mean shape of all the skulls 
plotRefToTarget(procrustes$consensus, mean_skull[,,1], gridPars=gridPar(pt.bg = "grey", pt.size = 0.5), method = "points", links= Wireframe, mag = 2)
```

#Import average species classifier
```{r}
Species_classifier<-read.csv('classifier_species.csv')
Species_classifier <- Species_classifier[match(dimnames(mean_skull)[[3]], Species_classifier$Species1),] #reorder the classifier to match shape data
csize <- Species_classifier$Csize; names(csize) <- Species_classifier$Species1 # give the size vector names so that it can be reordered later on

#Read linear measurements
BC_SL <- read.csv("Sp_Mean_BC_SL.csv", header = TRUE)
```

#PCA of species means
```{r}
PCA_species <- gm.prcomp(mean_skull)
plot(PCA_species)
#PC1 45.08, PC2 19.4
PCA_data$d #eigenvalues (not standardised)
sum(PCA_data$d)
EVALUES <- 100*(PCA_data$d)/sum(PCA_data$d) 
plot(EVALUES)   #scree plot
```

##PCA graph Colour by diet on mean species 
```{r}
mean_diet <- as.factor(Species_classifier$Diet) 
color2= rep(NA, length=length(Species_classifier$Diet))
color2[which(Species_classifier$Diet=="1")]= "brown"
color2[which(Species_classifier$Diet=="2")]= "green"
color2[which(Species_classifier$Diet=="3")]= "yellow"
color2[which(Species_classifier$Diet=="4")]= "blue"
color2[which(Species_classifier$Diet=="5")]= "red"
names(color2) <- Species_classifier$Species

PCA_mean_shape <- plot(PCA_species$x[,1], PCA_species$x[,2], bg = color2, col= "black", pch = 21, cex = csize/5, 
                     asp = F, xlab="Principal Component 1 (45.1%)", ylab="Principal Component 2 (19.4%)")
```

#Plotting the min and max of mean shape PC1 & PC2 to VISUALISE SHAPE CHANGE via VECTOR IMAGES
```{r}
plot.params <- gridPar(pt.bg="grey", pt.size=0.5, link.col="grey", tar.pt.bg = "black",tar.pt.size = 1,tar.link.col = "black",)
# pt.bg colour of dot
# pt.size size of dot
# link.col colour of wireframe
# link.lwd wireframe thickness
# link.lty wireframe line type (1 is solid. 2 dash etc.)

#Plot pc1 min and max against consensus 
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp1$max, method = "points", links = Wireframe, gridPars = plot.params)
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp1$min, method = "points", links = Wireframe, gridPars = plot.params)
#Plot pc2 min and max against consensus 
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp2$max, method = "points", links = Wireframe, gridPars = plot.params)
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp2$min, method = "points", links = Wireframe, gridPars = plot.params)
```

#ProcD.lm of size with coords, PC1 and PC2
```{r}
diet2 <- Species_classifier$Diet
names(diet2) <- Species_classifier$Species1

#Dataframe for procD
data_frame_species <- geomorph.data.frame(coords=mean_skull, PC1= PCA_species$x[,1], PC2= PCA_species$x[,2], size= csize, diet= diet2)

#Shape with size- is there allometry?
coords_csize <- procD.lm(coords~log(size), data=data_frame_species)
coords_csize$aov.table
#p=0.007 ; 20.9% variation due to allometry

#PC1 with size- was PC1 (main variation) driven by size? 
pc1_csize <- procD.lm(PC1~log(size), data=data_frame_species)
pc1_csize$aov.table
#p=0.190  ; 10% allometry
#Myoictis has elongate rostrum and incredibly flattened braincase from lateral view; rostrum and braincase widened; narrow zygomatic arches in dorsal view.

#PC2 with size- was PC2 driven by size?
pc2_csize <- procD.lm(PC2~log(size), data=data_frame_species)
pc2_csize$aov.table
#p=0.001  ; 80% allometry
```

#Plot shape and size regression graph 
```{r}
allom.plot <- plot(coords_csize, type = "regression", 
     predictor = log(data_frame_species$size), reg.type = "RegScore", 
     pch = 19, col = color2)
text(x=log(data_frame_species$size), y=allom.plot$RegScore, labels = dimnames(mean_skull)[[3]], pos=4, cex=0.5)
lines(x=log(data_frame_species$size), y=allom.plot$PredLine, pch=19, cex=0.5)
```

# ProcD.lm of diet with size, shape, pc1 and pc2 
```{r}
#PC1 with diet- Can PC1 separate the 5 diet categories?
pc1_diet <- procD.lm(PC1~diet, data=data_frame_species)
pc1_diet$aov.table
# p= 0.953

#PC2 with diet- Can PC2 separate the 5 diet categories?
pc2_diet <- procD.lm(PC2~diet, data=data_frame_species)
pc2_diet$aov.table
# p=0.001

#Is cranial shape significantly different between the 5 diet categories?
coords_diet <- procD.lm(coords~diet, data=data_frame_species)
coords_diet$aov.table 
# p=0.017

#Are the species in each of the 5 diet categories significantly different in size?
diet_csize <- procD.lm(diet~log(size), data = data_frame_species)
diet_csize$aov.table
# p=0.001, 80% allometry
```

#Import Phylogenetic tree
```{r}
###Tree with full names
###phylo_tree<- read.tree(text="(Antechinus_flavipes:17.07365857,((Sarcophilus_harrisii:7.12,(Dasyurus_hallucatus:6.425,(Dasyurus_maculatus:5.67801667,(Dasyurus_viverrinus:3.63,(Dasyurus_geoffroii:2.765,Dasyurus_albopunctatus:2.765):0.865):2.04801667):0.74698333):0.695):4.69486333,((Pseudantechinus_bilarni:9.63,(Pseudantechinus_woolleyae:8.97563333,(Pseudantechinus_macdonnellensis:8.4023,Pseudantechinus_ningbing:8.4023):0.57333333):0.65436667):0.645,(Parantechinus_apicalis:9.7,((Dasyuroides_byrnei:7.54,(Dasycercus_cristicauda:3.5,Dasycercus_blythi:3.5):4.04):1.73,(Myoictis_melas:8.1,Dasykaluta_rosamondae:8.1):1.17):0.43):0.575):1.53986333):5.25879524);")
##2021tree  
phylo_tree<- read.tree(text="(A.flavipes:17.07365857,((S.harrisii:7.12,(Das.hallucatus:6.425,(Das.maculatus:5.67801667,(Das.viverrinus:3.63,(Das.geoffroii:2.765,Das.albopunctatus:2.765):0.865):2.04801667):0.74698333):0.695):4.69486333,((Ps.bilarni:9.63,(Ps.woolleyae:8.97563333,(Ps.macdonnellensis:8.4023,Ps.ningbing:8.4023):0.57333333):0.65436667):0.645,(Pa.apicalis:9.7,((Dyu.byrnei:7.54,(Dyc.cristicauda:3.5,Dyc.blythi:3.5):4.04):1.73,(M.melas:8.1,Dyk.rosamondae:8.1):1.17):0.43):0.575):1.53986333):5.25879524);")
##2023tree with correct genus.species names 
#phylo_tree<- read.tree(text="(A.flavipes:18.03550000,(((((Pa.apicalis:11.28070000,M.melas:11.28070000)'14':3.09679000,Dyk.rosamondae:14.37749000)'13':0.30001000,(Dyu.byrnei:8.58000000,(Dyc.cristicauda:7.04618000,Dyc.blythi:7.04618000)'25':1.53382000)'37':6.09750000)'36':0.00000000,(Ps.bilarni:13.13022000,((Ps.macdonnellensis:8.72024000,Ps.ningbing:8.72024000)'35':1.11379000,Ps.woolleyae:9.83403000)'34':3.29619000)'43':1.54728000)'33':0.04382000,(S.harrisii:9.44000000,(Das.hallucatus:8.15000000,(((Das.geoffroii:3.43000000,Das.albopunctatus:3.43000000)'51':1.43000000,Das.viverrinus:4.86000000)'50':1.40702000,Das.maculatus:6.26702000)'49':1.88298000)'57':1.29000000)'60':5.28132000)'56':3.31418000);")
plot(phylo_tree)
```

# Reorder all data objects to match the order of tree tip labels
```{r}
phylo_data_frame_species <- data_frame_species
phylo_data_frame_species$coords <- phylo_data_frame_species$coords[,,match(phylo_tree$tip.label, dimnames(phylo_data_frame_species$coords)[[3]])]
phylo_data_frame_species$PC1 <- phylo_data_frame_species$PC1[match(phylo_tree$tip.label, names(phylo_data_frame_species$PC1))]
phylo_data_frame_species$PC2 <- phylo_data_frame_species$PC2[match(phylo_tree$tip.label, names(phylo_data_frame_species$PC2))]
phylo_data_frame_species$size <- phylo_data_frame_species$size[match(phylo_tree$tip.label, names(phylo_data_frame_species$size))]
phylo_data_frame_species$diet <- phylo_data_frame_species$diet[match(phylo_tree$tip.label, names(phylo_data_frame_species$diet))]
phylo_PCA_species <-PCA_species$x[match(phylo_tree$tip.label, rownames(PCA_species$x)),] #match the PCA data to the tree tip labels
```

# PGLS analyses of size on shape, PC1, PC2
```{r}
#PC1 with size- does size significantly drive the largest component of variation?
phylo_pc1_csize <- procD.pgls(PC1~log(size), data=phylo_data_frame_species, phy = phylo_tree)
phylo_pc1_csize$aov.table
plot (log(phylo_data_frame_species$size),phylo_data_frame_species$PC1)
#p=0.55; 

phylo_pc2_csize <- procD.pgls(PC2~log(size), data=phylo_data_frame_species, phy = phylo_tree)
phylo_pc2_csize$aov.table
plot (log(phylo_data_frame_species$size),phylo_data_frame_species$PC2)
#p=0.001  ; 67% allometry

phylo_coords_csize <- procD.pgls(coords~log(size), data=phylo_data_frame_species, phy = phylo_tree)
phylo_coords_csize$aov.table
#p=0.128 ; 10.8% variation due to allometry
```

# Graph regression plot (size and shape) (FIGURE 3B)
```{r}
phylo_mean_diet <- as.factor(phylo_data_frame_species$diet) 
color3= rep(NA, length=length(phylo_data_frame_species$diet))
color3[which(phylo_data_frame_species$diet=="1")]= "brown"
color3[which(phylo_data_frame_species$diet=="2")]= "green"
color3[which(phylo_data_frame_species$diet=="3")]= "yellow"
color3[which(phylo_data_frame_species$diet=="4")]= "blue"
color3[which(phylo_data_frame_species$diet=="5")]= "red"
names(color3) <- phylo_tree$tip.label

phylo_allom_plot <- plot(phylo_coords_csize, type = "regression", 
                   predictor = log(phylo_data_frame_species$size), reg.type = "RegScore", 
                   pch = 19, col =color3)
text(x=log(phylo_data_frame_species$size), y=phylo_allom_plot$RegScore, labels = dimnames(phylo_data_frame_species$coords)[[3]], pos=4, cex=0.5)
lines(x=log(phylo_data_frame_species$size), y=phylo_allom_plot$PredLine, pch=19, cex=0.5)
#phylomorphospace of the regression showing evolutionary allometry
#phylomorphospace(phylo_tree, cbind(log(phylo_data_frame_species$size),phylo_allom_plot$RegScore), label = c("horizontal"))
```

#PGLS- diet with PC1, PC2, size and size
```{r}
#PC1 with diet
phylo_pc1_diet <- procD.pgls(PC1~diet, data=phylo_data_frame_species, phy = phylo_tree)
phylo_pc1_diet$aov.table
#p= 0.322

phylo_pc2_diet <- procD.pgls(PC2~diet, data=phylo_data_frame_species, phy = phylo_tree)
phylo_pc2_diet$aov.table
#p=0.001

phylo_coords_diet <- procD.pgls(coords~diet, data=phylo_data_frame_species, phy = phylo_tree)
phylo_coords_diet$aov.table
#p=0.080

phylo_csize_diet <- procD.pgls(log(size)~diet, data=phylo_data_frame_species, phy = phylo_tree)
phylo_csize_diet$aov.table
# p=0.001, 61% allometry 
```

# Plot a phylomorphospace (FIGURE 3A)
```{r}
poly <- phylomorphospace(phylo_tree, phylo_PCA_species[,c(1:2)], label = c("horizontal"), node.size= c(0.4,1.5), xlab="Principal Component 1 (43.3%)", ylab="Principal Component 2 (20.0%)", fsize=0.8)
#poly.nolabels <- phylomorphospace(phylo_tree, PCA_species$x[,1:2], label = c("off"), node.size= c(0.5,1.5))
points(phylo_PCA_species[,c(1:2)], pch = 19, col =color3, cex=phylo_data_frame_species$size/5) #adds colour
```


#RESIDUALS: species coordinates that are predicted by CSIZE
```{r}
Presiduals <- coords_csize$residuals # procD.lm residuals

Presiduals <- Presiduals + coords_csize$fitted # Added the fitted to the residuals so that the resulting values look like skulls again

PresidualsArray <- arrayspecs(Presiduals,35,3) 
```

#PCA of species means of residuals 
#PCA graph of residuals (FIGURE 4)
#Vectors to visualise shape
```{r}
PCA_species_residuals <- gm.prcomp(PresidualsArray)
plot(PCA_species_residuals, pch=19, col=color2, cex=BC_SL$BasicranL/10)
text(x=PCA_species_residuals$x[,1], y=PCA_species_residuals$x[,2], labels = dimnames(mean_skull)[[3]], pos=4, cex=0.5)

summary(PCA_species_residuals)

#Plotting the min and max of PC1 to VISUALISE SHAPE CHANGE via VECTOR IMAGES
plotRefToTarget(procrustes$consensus, PCA_species_residuals$shapes$shapes.comp1$max, method = "vector")
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp1$min, method = "vector")
#plotting the min and max of PC2
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp2$max, method = "vector")
plotRefToTarget(procrustes$consensus, PCA_species$shapes$shapes.comp2$min, method = "vector")
```

#Plot Residual PC1 v Basicranial length 
```{r}
plot(y=PCA_species_residuals$x[,1], x=BC_SL$BasicranL, pch=19, col=color2, ylab="PC1 of shape residuals", xlab="Basicranial length")
text(y=PCA_species_residuals$x[,1], x=BC_SL$BasicranL, labels = dimnames(mean_skull)[[3]], pos=4, cex=0.5)
abline(h=0.05)
abline(h=-0.05)
```
